Type-checking Linearity in GHC Core

### 2024-01-24 icfp.tex
revised a bit of 2
Turns assumptions into lemmas, and make lemma 2 provable

Let bindings can also go unused if they are defined before branching on case
alternatives. At runtime, depending on the branch taken, the let binding will
be evaluated only if it occurs in that branch.
Both optimising transformations (float-out), and programmers used to non-strict evaluation, can produce
programs with bindings that are selectively used in the case alternatives, for
instance:
%
\begin{notyet}
\begin{lstlisting}
f3 :: (a *' $\lolli$ '* a) -> Bool -> a *' $\lolli$ '* a
f3 use bool x = let y = use x in case bool of
      True -> x 
      False -> y
\end{lstlisting}
\end{notyet}
%
This example essentially merges \lstinline{f1} with \lstinline{f2}, using \lstinline{x} directly in one
branch and using \lstinline{y} in the other. Semantically, this program is linear because
the linear resource \lstinline{x} ends up being used exactly once in both case
alternatives, directly or indirectly.

@@

>>>>

\todo[inline]{Cut this one}
Let bindings can also go unused if they are defined before branching on case
alternatives. At runtime, depending on the branch taken, the let binding will
be evaluated only if it occurs in that branch.
Both optimising transformations (float-out), and programmers used to non-strict evaluation, can produce
programs with bindings that are selectively used in the case alternatives, for
instance:
%
\begin{notyet}
\begin{lstlisting}
f3 :: (a *' $\lolli$ '* a) -> Bool -> a *' $\lolli$ '* a
f3 use bool x = let y = use x in case bool of
      True -> x 
      False -> y
\end{lstlisting}
\end{notyet}
%
This example essentially merges \lstinline{f1} with \lstinline{f2}, using \lstinline{x} directly in one
branch and using \lstinline{y} in the other. Semantically, this program is linear because
the linear resource \lstinline{x} ends up being used exactly once in both case
alternatives, directly or indirectly.

@@

\renewcommand{\DeltaLinearRelationLemma}{
\begin{lemma}[$\D \Rightarrow 1$]
Any $\Delta$-variable with a fully-irrelevant usage environment can be made a linear variable as long as its irrelevant environment is dropped from the linear environment (and the unrestricted env. is updated accordingly).\\
If $\G,\x[\irr{\D}]; \irr{\D},\D' \vdash e : \vp$
then $\G[x/\irr{\D}]; \D',\xl \vdash e :\vp$.
\end{lemma}
}

\renewcommand{\LinearDeltaRelationLemma}{
\begin{lemma}[$1 \Rightarrow \D$]
A linear variable can be moved to the unrestricted context as a $\D$-var with any usage environment $\D$ by introducing $\D$ in the linear resources (and the unrestricted env. is updated accordingly).\\
If $\G; \D',\xl \vdash e :\vp$
then $\G[\D/x],\xD; \D,\D' \vdash e : \vp$.
\end{lemma}
}

\renewcommand{\DeltaUnrestrictedRelationLemma}{
\begin{lemma}[$x{:}_\omega\sigma = x{:}_{\cdot}\sigma$]
An unrestricted variable is equivalent to a $\D$-var with an empty usage environment.\\
$\G,\xo; \D \vdash e : \vp$ iff $\G,\x[\cdot]; \D \vdash e : \vp$
\end{lemma}
}
